---
title: 'Wrangling Data from microplate OD600 growth curve in 96WP'
author: "El Park"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

R Setup
```{r setup}
rm(list = ls())
knitr::opts_knit$set(root.dir = "C:/Users/parke/OneDrive - Indiana University/GitHub/ASG-fitness-effects/microplate_fitness_assay/")
getwd()

#install.packages("tidyverse")
require("png")
require("dplyr")
library("tidyverse")
require("grid")
require("tibble")
require("knitr")
require("extrafont")
require("ggrepel");
require("gridExtra")
require("contrast")
sem <- function(x) sqrt(var(x)/length(x))
cv <- function(x) (sd(x)/mean(x))*100

```

Load and wrangle data
```{r}
#load data
data<-read.csv("C:/Users/parke/OneDrive - Indiana University/GitHub/ASG-fitness-effects/microplate_fitness_assay/data/20250204OD600.csv")
design<-read.csv("C:/Users/parke/OneDrive - Indiana University/GitHub/ASG-fitness-effects/microplate_fitness_assay/reference/20250204OD600Design.csv", header=FALSE)

#subset design file, "wells" will be strain plus treatment but I can update design file to include a separate column for both
#media<-design[1:8,1:12] all same media
wells<-design[9:16,1:12]

# Create custom labels
rows <- LETTERS[1:8]  # A to H
cols <- 1:12  # 1 to 12
custom_labels <- unlist(lapply(rows, function(row) paste0(row, cols)))

#convert to long
#media.long <- as.data.frame(pivot_longer(media, cols = everything(),
                                         #names_to = "Treatment", values_to = "Value"))
wells.long <- as.data.frame(pivot_longer(wells, cols = everything(),
                                         names_to = "Treatment", values_to = "Value"))

# Replace the "Treatment" column values
#media.long<-media.long %>%
    #mutate(Treatment = custom_labels[1:n()])
wells.long<-wells.long %>%
    mutate(Treatment = custom_labels[1:n()])

#colnames(media.long)<-c("Well", "Media") #to rename every column
colnames(wells.long)<-c("Well", "Treatment") #to rename every column


# Prepare the OD data for merging
OD.long <- data %>%
  pivot_longer(cols = -Time..Hours., names_to="Well", values_to="OD600")
str(OD.long)

#combine
#OD.long$Media<-rep(media.long$Media, length.out=18528)
OD.long$Treatment<-rep(wells.long$Treatment, length.out=18528)

#filter out "Blanks" (what if I made them na and did na.omit, would that be easier?)
summary<-OD.long[!apply(OD.long, 1, function(row) any(row=="Blank")), ]
summary<-summary[!apply(summary, 1, function(row) any(row=="Ctrl")), ]

#it worked!!
```

Process data (infected cells)
#I need to create two vectors for each medium:
##OD over time for wt during infection
##OD over time for mut during infection
#Then I need to take the mean of the wt replicates for each time point and repeat
#I can subtract mut from wt(means) to get a number corresponding to "fitness"
```{r}
#Make a matrix that includes all infected 
SPO1<-summary %>%
  filter(grepl("SPO1", Treatment)) %>%
  select(Time..Hours., OD600, Treatment)
#subset wt
wt<-SPO1 %>%
  filter(grepl("WT", Treatment)) %>%
  select(Time..Hours., OD600, Treatment)

#take averages
wtmeans <- wt %>%
  group_by(Time..Hours.) %>%
  summarise(mean_OD600 = mean(OD600))

# Repeat each row 4 times (n=4 in this specific exp)
wtmeans <-wtmeans[rep(1:nrow(wtmeans), each = 6),]


#now SPOIIE
SPOIIE<-SPO1 %>%
  filter(grepl("SPOIIE", Treatment)) %>%
  select(Time..Hours., OD600)



#subract SPOIIE from WT and then take sum
fitness<-SPOIIE-wtmeans
fitness<-print(sum(fitness))




```
Process data (noninfected cells)
#I need to create two vectors for each medium:
##OD over time for wt during infection
##OD over time for mut during infection
#Then I need to take the mean of the wt replicates for each time point and repeat
#I can subtract mut from wt(means) to get a number corresponding to "fitness"
```{r}
#Make a matrix that includes all infected 
noninf<-summary %>%
  filter(!grepl("SPO1", Treatment)) %>%
  select(Time..Hours., OD600, Treatment)
#subset wt
n.wt<-noninf %>%
  filter(grepl("WT", Treatment)) %>%
  select(Time..Hours., OD600)
#take averages
n.wtmeans <- n.wt %>%
  group_by(Time..Hours.) %>%
  summarise(mean_OD600 = mean(OD600))
# Repeat each row 4 times (n=4 in this specific exp)
n.wtmeans <-n.wtmeans[rep(1:nrow(n.wtmeans), each = 6),]

#now SPOIIE
n.SPOIIE<-noninf %>%
  filter(grepl("SPOIIE", Treatment)) %>%
  select(Time..Hours., OD600)

#subract SPOIIE from WT and then take sum
n.fitness<-n.SPOIIE-n.wtmeans
n.fitness<-print(sum(n.fitness))

```

Add Strain and Infection columns
```{r}
summary <- summary %>%
  filter(Treatment != "Ctrl") %>%
  mutate(SPO1_Presence = ifelse(grepl("SPO1", Treatment), "Infected", "Noninfected"))
summary <- summary %>%
  mutate(Strain = ifelse(grepl("WT", Treatment), "WT", "SPOIIE"))
```

Plot data
```{r}
setwd('./output') #for ggsave
plot.new()

# Define plotting margins
par(mar = c(7, 7, 5, 7))

#vector of all the fitness values (Fitness effect of loss of gene)
pm <-data.frame(
  treatment_media = c("noninfected", "infected"),
  fitness = c(n.fitness, fitness)
)



bg<-ggplot(pm, aes(x = treatment_media, y = fitness)) +
  geom_bar(stat = "identity") +
  labs(title = "Fitness effects of sporulation mutations", x = "Treatment and Media", y = "Fitness") +
  theme(legend.key.size = unit(1, 'cm'), legend.position = "right")
ggsave("20250204_fitness.png", plot = bg, width = 6, height = 4, dpi = 300)
print(bg)

# Plot all growth over time in one plot, colored by design
p<-ggplot(summary, aes(x = Time..Hours., y = OD600, colour = Treatment))+
  geom_line() +
  theme_minimal() + geom_smooth()+
  labs(title = "OD600 over Time by Cell Type", x = "Time", y = "OD600", colour = "Strain-Treatment") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right")+
  facet_wrap(~Treatment)
print(p)
ggsave("20250204_all.png", plot = p, width = 12, height = 6, dpi = 300)

#plot and sort by strain
d<-ggplot(summary, aes(x = Time..Hours., y = OD600, colour = SPO1_Presence)) +
  #geom_line() +
  theme_minimal() + geom_smooth()+
  labs(title = "OD600 over Time by Cell Type and Treatment", x = "Time", y = "OD600", colour = "SPO1_Presence") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right")+
  facet_wrap(~Strain)
print(d)
ggsave("20250204_Strains.png", plot = d, width = 6, height = 4, dpi = 300)

#plot and sort by infection
s <- ggplot(summary, aes(x = Time..Hours., y = OD600, colour = Strain)) +
  theme_minimal() + 
  geom_smooth() +
  labs(title = "OD600 over Time by Cell Type and Treatment", x = "Time", y = "OD600", colour = "Strain") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right") +
  facet_wrap(~SPO1_Presence)
print(s)
ggsave("20250204_Infection.png", plot = s, width = 6, height = 4, dpi = 300)

```
Results:
>
