---
title: 'Wrangling Data from microplate OD600 growth curve in 96WP'
author: "El Park"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

R Setup
```{r setup}
rm(list = ls())
setwd("./")
getwd()

#install.packages("tidyverse")
require("png")
require("dplyr")
library("tidyverse")
require("grid")
require("tibble")
require("knitr")
require("extrafont")
require("ggrepel");
require("gridExtra")
require("contrast")
sem <- function(x) sqrt(var(x)/length(x))
cv <- function(x) (sd(x)/mean(x))*100

```

Load and wrangle data
```{r}
#load data
data<-read.csv("../data/20250228OD600.csv")
design<-read.csv("../reference/20250228OD600Design.csv", header=FALSE)

#subset design file, "wells" will be strain plus treatment but I can update design file to include a separate column for both
media<-design[1:8,1:12]
strain<-design[9:16,1:12]
infection<-design[17:24, 1:12]


# Create custom labels
rows <- LETTERS[1:8]  # A to H
cols <- 1:12  # 1 to 12
custom_labels <- unlist(lapply(rows, function(row) paste0(row, cols)))

#convert to long
med.long <- as.data.frame(pivot_longer(media, cols = everything(),
                                         names_to = "Well", values_to = "Media"))
stn.long <- as.data.frame(pivot_longer(strain, cols = everything(),
                                         names_to = "Well", values_to = "Strain"))
inf.long <- as.data.frame(pivot_longer(infection, cols = everything(),
                                         names_to = "Well", values_to = "Infection"))

# Prepare the OD data for merging
OD.long <- data %>%
  pivot_longer(cols = -Time, names_to="Well", values_to="OD600")
str(OD.long)

#combine
OD.long$Media<-rep(med.long$Media, length.out=55392)
OD.long$Strain<-rep(stn.long$Strain, length.out=55392)
OD.long$Infection<-rep(inf.long$Infection, length.out=55392)


#filter out "Blanks" 
summary<-OD.long[!apply(OD.long, 1, function(row) any(row=="Blank")), ]

#it worked!!
```

Calculate fitness
```{r}
# Get unique combinations of Media, Infection, and Strain
combinations <- summary %>%
  distinct(Media, Infection, Strain)

# Create data frames for each combination
for (i in 1:nrow(combinations)) {
  strain <- combinations$Strain[i]
  media <- combinations$Media[i]
  infection <- combinations$Infection[i]
  
  vector_name <- paste(strain, media, infection, sep = ".")
  assign(vector_name, summary %>%
    filter(Strain == strain, Media == media, Infection == infection) %>%
    select(Time, OD600))
}

# Calculate the average OD600 for each time point for wt strains
wt_combinations <- combinations %>%
  filter(grepl("WT", Strain))

for (i in 1:nrow(wt_combinations)) {
  media <- wt_combinations$Media[i]
  infection <- wt_combinations$Infection[i]
  
  wt_vector_name <- paste("WT", media, infection, sep = ".")
  avg_vector_name <- paste("avg", media, infection, sep = ".")
  
  avg_vector <- get(wt_vector_name) %>%
    group_by(Time) %>%
    summarise(avg_OD600 = mean(OD600, na.rm = TRUE)) %>%
    mutate(avg_OD600 = avg_OD600 * 3)
  
  assign(avg_vector_name, avg_vector)
}

# Ensure non_wt_combinations is created correctly
non_wt_combinations <- combinations %>%
  filter(!grepl("WT", Strain))

fitness_list <- list()

for (i in 1:nrow(non_wt_combinations)) {
  strain <- non_wt_combinations$Strain[i]
  media <- non_wt_combinations$Media[i]
  infection <- non_wt_combinations$Infection[i]
  
  non_wt_vector_name <- paste(strain, media, infection, sep = ".")
  avg_vector_name <- paste("avg", media, infection, sep = ".")
  fitness_vector_name <- paste(media, strain, infection, "fitness", sep = ".")
  
  fitness_vector <- get(non_wt_vector_name)$OD600 - get(avg_vector_name)$avg_OD600
  
  assign(fitness_vector_name, fitness_vector)
  fitness_list[[fitness_vector_name]] <- fitness_vector
}

# Sum all fitness vectors into one data frame named pm
pm <- data.frame(
  media = character(),
  strain = character(),
  infection = character(),
  fitness_value = numeric()
)

for (name in names(fitness_list)) {
  parts <- strsplit(name, "\\.")[[1]]
  media <- parts[1]
  strain <- parts[2]
  infection <- parts[3]
  fitness_value <- sum(fitness_list[[name]])
  
  pm <- rbind(pm, data.frame(media = media, strain = strain, infection = infection, fitness_value = fitness_value))
}

# Create the media.strain.infection variable manually
pm$media.strain.infection <- paste(pm$media, pm$strain, pm$infection, sep = ".")


# Plot fitness
plot.new()
bg <- ggplot(pm, aes(x = media.strain.infection, y = fitness_value)) +
  geom_bar(stat = "identity") +
  labs(title = "Fitness effects of sporulation mutations", x = "Media.Strain.Infection", y = "Fitness") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8), # Rotate and adjust size of x-axis labels
        legend.key.size = unit(1, 'cm'), 
        legend.position = "right") +
  scale_y_continuous(breaks = c(pretty(pm$fitness_value), 300)) # Add tick mark at 500
print(bg)

```


#subset wt
wt<-summary %>%
  filter(grepl("WT", Strain)) %>%
  select(Time, OD600, Media, Infection)
#separate by media and infection (ctrl)
wtLB<-wt %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("Ctrl", Infection)) %>%
  select(Time, OD600)
wtDSM<-wt %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("Ctrl", Infection)) %>%
  select(Time, OD600)
#separate by media and infection (SPO1)
wtlbSPO1<-wt %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time, OD600)
wtdsmSPO1<-wt %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time, OD600)

#separate by media and infection (sspSPO1)
wtlbsspSPO1<-wt %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("sspSPO1", Infection)) %>%
  select(Time, OD600)
wtdsmsspSPO1<-wt %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("sspSPO1", Infection)) %>%
  select(Time, OD600)

#separate by media and infection (SPP1)
wtlbSPP1<-wt %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("SPP1", Infection)) %>%
  select(Time, OD600)
wtdsmSPP1<-wt %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("SPP1", Infection)) %>%
  select(Time, OD600)

#take averages
wtLBmeans <- wtLB %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtDSMmeans <- wtDSM %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtlbSPO1means <- wtlbSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtdsmSPO1means <- wtdsmSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtlbsspSPO1means <- wtlbSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtdsmsspSPO1means <- wtdsmSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtlbSPP1means <- wtlbSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))
wtdsmSPP1means <- wtdsmSPO1 %>%
  group_by(Time) %>%
  summarise(mean_OD600 = mean(OD600))

# Repeat each row n times (n=4 in this specific exp)
wtLBmeans <-wtLBmeans[rep(1:nrow(wtLBmeans), each = 4),]
wtDSMmeans <- wtDSMmeans[rep(1:nrow(wtDSMmeans), each = 4),]
wtlbSPO1means <-wtlbSPO1means[rep(1:nrow(wtlbSPO1means), each = 4),]
wtdsmSPO1means <- wtdsmSPO1means[rep(1:nrow(wtdsmSPO1means), each = 4),]
wtlbsspSPO1means <-wtlbsspSPO1means[rep(1:nrow(wtlbsspSPO1means), each = 4),]
wtdsmsspSPO1means <- wtdsmsspSPO1means[rep(1:nrow(wtdsmsspSPO1means), each = 4),]
wtlbSPP1means <-wtlbSPP1means[rep(1:nrow(wtlbSPP1means), each = 4),]
wtdsmSPP1means <- wtdsmSPP1means[rep(1:nrow(wtdsmSPP1means), each = 4),]

#now SPOIIE
SPOIIE<-summary %>%
  filter(grepl("SPOIIE", Strain)) %>%
  select(Time..Hours., OD600, Media, Infection)
#separate by media and infection (ctrl)
SPOIIELB<-SPOIIE %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("Ctrl", Infection)) %>%
  select(Time..Hours., OD600)
SPOIIEDSM<-SPOIIE %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("Ctrl", Infection)) %>%
  select(Time..Hours., OD600)
#separate by media and infection (SPO1)
SPOIIElbSPO1<-SPOIIE %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time..Hours., OD600)
SPOIIEdsmSPO1<-SPOIIE %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time..Hours., OD600)
#separate by media and infection (SPO1)
SPOIIElbSPO1<-SPOIIE %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time..Hours., OD600)
SPOIIEdsmSPO1<-SPOIIE %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("SPO1", Infection)) %>%
  select(Time..Hours., OD600)
#separate by media and infection (SPO1)
SPOIIElbSPP1<-SPOIIE %>%
  filter(grepl("LB", Media)) %>%
  filter(grepl("SPP1", Infection)) %>%
  select(Time..Hours., OD600)
SPOIIEdsmSPP1<-SPOIIE %>%
  filter(grepl("DSM", Media)) %>%
  filter(grepl("SPP1", Infection)) %>%
  select(Time, OD600)


#subract SPOIIE from WT and then take sum
fLB<-(SPOIIELB-wtLBmeans)
fitnessLB<-sum(fLB)
flbSPO1<-(SPOIIElbSPO1-wtlbSPO1means)
fitnesslbSPO1<-sum(flbSPO1)

fDSM<-SPOIIEDSM-wtDSMmeans
fitnessDSM<-sum(fDSM)
fdsmSPO1<-SPOIIEdsmSPO1-wtdsmSPO1means
fitnessdsmSPO1<-sum(fdsmSPO1)

print(done)



Plot data
```{r}
setwd('C:/Users/parke/OneDrive - Indiana University/GitHub/ASG-fitness-effects/microplate_fitness_assay/output') #for ggsave

# Plot all growth over time in one plot, colored by design
p<-ggplot(summary, aes(x = Time, y = OD600, colour = interaction(Strain, Media, Infection)))+
  geom_smooth() +
  theme_minimal() + geom_smooth()+
  labs(title = "OD600 over Time for All Samples", x = "Time", y = "OD600", colour = "Media.Strain.Treatment") +
  theme(legend.key.size = unit(1, 'cm'), legend.position = "right")+
  facet_wrap(~Strain + Media + Infection)
ggsave("20250228_all.png", plot = p, width = 10, height = 12, dpi = 300)
print(p)

#plot and sort by Media
m<-ggplot(summary, aes(x = Time, y = OD600, colour = interaction(Strain, Infection)))+
  #geom_line() + 
  theme_minimal() + geom_smooth()+
  labs(title = "OD600 over Time by Media", x = "Time", y = "OD600", colour = "Strain.Treatment") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right")+
  facet_wrap(~Media)
ggsave("20250228_Media.png", plot = m, width = 8, height = 5, dpi = 300)
print(m)

#plot and sort by treatment
d<-ggplot(summary, aes(x = Time, y = OD600, colour = interaction(Media, Infection))) +
  #geom_line() +
  theme_minimal() + geom_smooth()+
  labs(title = "OD600 over Time by Strain and Media", x = "Time", y = "OD600", colour = "Media.Infection") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right")+
  facet_wrap(~Strain+Media)
ggsave("20250228_Strains.png", plot = d, width = 8, height = 6, dpi = 300)
print(d)

#plot and sort by infection
s <- ggplot(summary, aes(x = Time..Hours., y = OD600, colour = interaction(Strain, Media))) +
  theme_minimal() + 
  geom_smooth() +
  labs(title = "OD600 over Time by Treatment", x = "Time", y = "OD600", colour = "Strain") +
  theme(legend.key.size = unit(0.5, 'cm'), legend.position = "right") +
  facet_wrap(~Infection)
ggsave("20250228_Infection.png", plot = d, width = 6, height = 4, dpi = 300)
print(s)

```
Results:
>
